---
title: Introduction to R
author: Naom N
output: pdf_document
date: "2023-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question one: Nuts and Bolts of R
a.) Using seq() function, create a sequence from -2.0 to 2.0 with increments of 0.25. What are the sample mean and standard deviation of theses numbers?

```{r}
a<-seq(-2,2,0.25)
mean(a)
sd(a)
```
b.) Using seq() function, create a sequence from -4.0 to 4.0 with increments of 0.5. What are the sample mean and standard deviation of theses numbers?

```{r}
b<-seq(-4,4,0.5)
mean(b)
sd(b)
```
c.) Compare the means and standard deviations from part (a) and (b). What do you conclude and explain why.

Answer:
The means are the same but the standard deviations are not.
Since the standard deviation measures dispersion from the mean, the second sequence will have a higher deviation as the numbers are spread out far from 0 compared to the first sequence.


d.) Place the two sequences from parts a and b into a 17 × 2 matrix as rows and find the sum of each row of the matrix. Try both for loop and apply()

```{r}
c<-matrix(c(a,b),ncol=2)

for (i in 1:17) {
  print(sum(c[i,]))
}

apply(c, 1, sum)
```

##Question two: The standard Normal distribution
a.) Write a function whose arguments are a vector and a number. The function should output the proportion of values in the vector less than the number. Print your function code here
```{r}
func=function(d,e){
  mean(d<e)
}
```

b.) Take 10,000 draws from the normal distribution with mean five and standard deviation one. (Hint: Use the function, rnorm(), to do this.) Store the 10,000 draws in a vector, but do not print the vector here. Print the mean and standard deviation of the numbers. Also make a histogram of the numbers and have a blue dashed density line fitted above.(Hint: Check the help file of lines function.)

```{r}
draws<-rnorm(10000,5,1)
mean(draws)
sd(draws)
hist(draws, probability = TRUE)
lines(density(draws),col='blue')
```
c.) Using your function from part (a) and your vector from part (b), what is an estimate of the probability that a normal random variable is less than 4 to be?

```{r}
func(draws,4)
```

d.) How do you expect the true probability that this random variable is greater than 6 to compare to the true probability that it is less than 4? Give evidence of this using your histogram

Answer:
The two probabilities will be close to each other
The histogram is almost perfectly symmetrical and we can see that the number probability
on the y-axis corresponding to both the region from 0-4 and 6-10 is between 0.1 and 0.2


##Question Three: Rock-paper-scissors
a.) Set up the rock paper scissors game using the following, and paste your code as the answer to this part.
Make sure you have the seed set so that you can reproduce the results.
• Specify a vector, x, of length 10 containing a random sample of the words ‘Rock’, ‘Paper’, or ‘Scissors’.
• Specify another such vector, y, of length 10 containing a random sample of the words ‘Rock’, ‘Paper’, or ‘Scissors’.
• Create a 2 × 10 matrix with x in the first row and y in the second row. Output this matrix.
• Create a vector of length 10 called ‘win’. In each element of the vector, record an ‘x’ if player ‘x’ won, a ‘y’ if player ‘y’ won, and a ‘d’ if the players drew (tied). Do this by writing a for loop with if...else... statement. Output win.

```{r}
set.seed(1)
n<-10
X<-sample(c("rock","paper","scissors"),n,replace = TRUE)
Y<-sample(c("rock","paper","scissors"),n,replace = TRUE)
rps_game<-as.matrix(rbind(X,Y))
rps_game

win<-rep(NA,n)
for (i in 1:length(win)) {
win[i]=(  if(rps_game[1,i] == rps_game[2,i]){("d")
  }
  else if(rps_game[1,i]=="rock" & rps_game[2,i]=="scissors" | rps_game[1,i]=="scissors"
          & rps_game[2,i]=="paper" |rps_game[1,i]=="paper" & rps_game[2,i]=="rock" ){
    ("x")
  }
  else{
    ("y")
  })
}
win
```

b.) Using your vector called win, what are estimates for: the probability that player ‘x’ won, the probability that player ‘y’ won, and the probability that the players tied?

```{r}

probs<-matrix(c(x_win<-mean(win=="x"),y_win<-mean(win=="y"),drew<-mean(win=="d")),
               nrow=3,ncol=1)
rownames(probs)<-c("X_prob","Y_prob","Draw")
probs
```

c.) Run your code again for a larger number of trials (i.e., for vector longer than length 10). Do this by writing a function f.x with number of trials as input and probabilities of winning as outputs. Try 20, 50 and 100 trials with f.x.

```{r}
set.seed(2)
f.x=function(g){
X<-sample(c("rock","paper","scissors"),g,replace = TRUE)
Y<-sample(c("rock","paper","scissors"),g,replace = TRUE)
rps_game<-as.matrix(rbind(X,Y))
rps_game

win<-rep(NA,g)
for (i in 1:length(win)) {
win[i]=(  if(rps_game[1,i] == rps_game[2,i]){("d")
  }
  else if(rps_game[1,i]=="rock" & rps_game[2,i]=="scissors" | rps_game[1,i]=="scissors"
          & rps_game[2,i]=="paper" |rps_game[1,i]=="paper" & rps_game[2,i]=="rock" ){
    ("x")
  }
  else{
    ("y")
  })
}
probs<-matrix(c(x_win<-mean(win=="x"),y_win<-mean(win=="y"),drew<-mean(win=="d")),
               nrow=3,ncol=1)
rownames(probs)<-c("X_prob","Y_prob","Draw");probs
}
f.x(20)
f.x(50)
f.x(100)
```

d.) What do you notice about the probability estimates from parts (b) and (c)?

Answer:
The probabilities are slowly getting close to each other as the number of trials increases.

