---
title: "Summarizing data - bingo game & cleaning data"
author: "Naom Nyambariga"
date: "2023-02-21"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##QUESTION 1
#Bingo
a). Randomly generate a classic 5*5 Bingo card of numbers


```{r}
# generate columns
  col1 <- sample(1:15, 5, replace = F)
  col2 <- sample(16:30, 5, replace = F)
  col3 <- sample(31:45, 5, replace = F)
  col4 <- sample(46:60, 5, replace = F)
  col5 <- sample(61:75, 5, replace = F)
  
  # append columns to create the bingo board 
  bingo_board<-c(col1,col2,col3,col4,col5)
```

b). Randomly select numbers from 1 to 75 without replacement, and “mark” the corresponding number on your card computationally however you choose.In total, 35 numbers will be randomly selected

```{r}
round_answers <- sample(1:75, 35, replace = F)
```


c). As soon as all the numbers in a row or column or either diagonal have been marked, the game ends and you win. Create a result object that tells whether you won by row, column, or diagonal
```{r}

  # generate bingo results for the current round
  round_answers <- sample(1:75, 35, replace = F)

  # mark the winning bingo cells in 5x5 grid
  for (j in 1:25) {
    bingo_board[j] = (
      if ((bingo_board[j] %in% round_answers) == TRUE) {
        0
      }
      else{
        1
      }
    )
  }
  bingo_ans_mat <- matrix(bingo_board, 5, 5)
  bingo_ans_mat[3, 3] <- 0
  
win_type <- "No winner"
  # verify the direction of the winning bingo set
  for (k in 1:5) {
    # check if there is winner by  row
    if (sum(c(bingo_ans_mat[k, ])) == 0) {
      win_type <- "won by row"
      break
    }
    # check if there is winner by  column
    if (sum(c(bingo_ans_mat[ ,k])) == 0) {
      win_type <- "won by column"
      break
    }
  }
  
  # check if there is a diagonal winner
  if (sum(c(
    bingo_ans_mat[1, 1],
    bingo_ans_mat[2, 2],
    bingo_ans_mat[3, 3],
    bingo_ans_mat[4, 4],
    bingo_ans_mat[5, 5]
  )) ==
  0 |
  sum(c(
    bingo_ans_mat[5, 1],
    bingo_ans_mat[4, 2],
    bingo_ans_mat[3, 3],
    bingo_ans_mat[2, 4],
    bingo_ans_mat[1, 5]
  )) ==
  0) {
    win_type <- "won by diagonal"
    }
```



d). Using this setup, play the game 25 times to estimate the probability of winning by marking a row, a column, or a diagonal set of spaces

```{r}
    bingo_game<-function(num_iterations,n) {
    experiment_win<-c()
    for (i in 1:num_iterations) {
      
      total_wins <-c()
      for (i in 1:n) {
        
        # generate columns
        col1 <- sample(1:15, 5, replace = F)
        col2 <- sample(16:30, 5, replace = F)
        col3 <- sample(31:45, 5, replace = F)
        col4 <- sample(46:60, 5, replace = F)
        col5 <- sample(61:75, 5, replace = F)
        
        # append columns to create the bingo board 
        bingo_board<-c(col1,col2,col3,col4,col5)
        
        # generate bingo results for the current round
        round_answers <- sample(1:75, 35, replace = F)
        
        # mark the winning bingo cells in 5x5 grid
        for (j in 1:25) {
          bingo_board[j] = (
            if ((bingo_board[j] %in% round_answers) == TRUE) {
              0
            }
            else{
              1
            }
          )
        }
        bingo_ans_mat <- matrix(bingo_board, 5, 5)
        bingo_ans_mat[3, 3] <- 0
        
        win_type <- "No winner"
        # verify the direction of the winning bingo set
        for (k in 1:5) {
          # check if there is winner by  row
          if (sum(c(bingo_ans_mat[k, ])) == 0) {
            win_type <- "won by row"
            break
          }
          # check if there is winner by  column
          if (sum(c(bingo_ans_mat[ ,k])) == 0) {
            win_type <- "won by column"
            break
          }
        }
        
        # check if there is a diagonal winner
        if (sum(c(
          bingo_ans_mat[1, 1],
          bingo_ans_mat[2, 2],
          bingo_ans_mat[3, 3],
          bingo_ans_mat[4, 4],
          bingo_ans_mat[5, 5]
        )) ==
        0 |
        sum(c(
          bingo_ans_mat[5, 1],
          bingo_ans_mat[4, 2],
          bingo_ans_mat[3, 3],
          bingo_ans_mat[2, 4],
          bingo_ans_mat[1, 5]
        )) ==
        0) {
          win_type <- "won by diagonal"
        }
        
        # append the winning bingo type to out total_wins vector
        total_wins <- append(total_wins, win_type)
      }
      total_wins
      
      #append total wins to experiment_win vector
      experiment_win<-append(experiment_win, total_wins)
    }
    experiment_win
    }
    
    bingo_game(1,25)
    
```
i).  A table with probability estimation for winning by row, column, diagonal and probability of fail for 10 experiments

```{r}
experiments_10<-bingo_game(10,25)
prop.table(table(experiments_10))
```
ii). A histogram for the winning probability among 100 experiments (no need to be categorized by row/column or diagonal).

```{r}
#where 0 represents no winner and 1 a winner

experiments_100 <- bingo_game(100,25)
numeric_wins <- 1 * (experiments_100 != "No winner")
hist(numeric_wins,  breaks = 2, freq = FALSE)

```
e). Using the same setup, play the game 250 times to estimate the probability of winning by marking a row, a column, or a diagonal set of spaces

i.) a table with probability estimation for winning by row, column, diagonal and probability of fail for 10 experiments

```{r}
experiments2_10<-bingo_game(10,250)
prop.table(table(experiments2_10))

```
ii). a histogram for the winning probability among 100 experiments (no need to be categorized by row/column or diagonal)

```{r}
#where 0 represents no winner and 1 a winner

experiments2_100 <- bingo_game(100,250)
numeric_wins2 <- 1 * (experiments2_100 != "No winner")
hist(numeric_wins2,  breaks = 2, freq = FALSE)
```
f). In another variation of Bingo, a wining card also can occur when each of the four corners is marked

i.) your bingo function

```{r}
  bingo_game2<-function(num_iterations,n) {
    experiment_win<-c()
    for (i in 1:num_iterations) {
      
      total_wins <-c()
      for (i in 1:n) {
        
        # generate columns
        col1 <- sample(1:15, 5, replace = F)
        col2 <- sample(16:30, 5, replace = F)
        col3 <- sample(31:45, 5, replace = F)
        col4 <- sample(46:60, 5, replace = F)
        col5 <- sample(61:75, 5, replace = F)
        
        # append columns to create the bingo board 
        bingo_board<-c(col1,col2,col3,col4,col5)
        
        # generate bingo results for the current round
        round_answers <- sample(1:75, 35, replace = F)
        
        # mark the winning bingo cells in 5x5 grid
        for (j in 1:25) {
          bingo_board[j] = (
            if ((bingo_board[j] %in% round_answers) == TRUE) {
              0
            }
            else{
              1
            }
          )
        }
        bingo_ans_mat <- matrix(bingo_board, 5, 5)
        bingo_ans_mat[3, 3] <- 0
        
        win_type <- "No winner"
        # verify the direction of the winning bingo set
        for (k in 1:5) {
          # check if there is winner by  row
          if (sum(c(bingo_ans_mat[k, ])) == 0) {
            win_type <- "won by row"
            break
          }
          # check if there is winner by  column
          if (sum(c(bingo_ans_mat[ ,k])) == 0) {
            win_type <- "won by column"
            break
          }
        }
        
        # check if there is a diagonal winner
        if (sum(c(
          bingo_ans_mat[1, 1],
          bingo_ans_mat[2, 2],
          bingo_ans_mat[3, 3],
          bingo_ans_mat[4, 4],
          bingo_ans_mat[5, 5]
        )) ==
        0 |
        sum(c(
          bingo_ans_mat[5, 1],
          bingo_ans_mat[4, 2],
          bingo_ans_mat[3, 3],
          bingo_ans_mat[2, 4],
          bingo_ans_mat[1, 5]
        )) ==
        0) {
          win_type <- "won by diagonal"
        }
        
        #check if there is a corner win
        if(sum(c(
          bingo_ans_mat[1, 1],
          bingo_ans_mat[1, 5],
          bingo_ans_mat[5, 1],
          bingo_ans_mat[5, 5]
        ))==
        0){
          win_type<-"won by corner"
        }
        
        # append the winning bingo type to out total_wins vector
        total_wins <- append(total_wins, win_type)
      }
      total_wins
      
      #append total wins to experiment_win vector
      experiment_win<-append(experiment_win, total_wins)
    }
    experiment_win
    }
    
#to check if there is a corner win in 20 games (check code)
    bingo_game2(1,20)
```

ii). assume the game is repeated for 250 times, make a table with probability estimation for winning by row, column, diagonal, corner and probability of fail for 10 experiments

```{r}
experiments3_10<-bingo_game2(10,250)
prop.table(table(experiments3_10))
```

iii). a histogram for the winning probability among 100 experiments (no need to be categorized by row/column, diagonal or corner

```{r}
#where 0 represents no winner and 1 a winner

experiments3_100 <- bingo_game(100,250)
numeric_wins3 <- 1 * (experiments3_100 != "No winner")
hist(numeric_wins3,  breaks = 2, freq = FALSE)
```

iv). illustrate your findings comparing the table and histogram with part (e)
Answer: 
The probabilities of there being no winner are very close to each other at around 72%
i,e 0.7188 for part e and 0.7204 for part f. Whether a winning type is added or not, the probability of not winning is pretty cnsistent. This is supported by the two histograms

##QUESTION TWO
#Diamond data set

a). Download and import the diamond data set from Blackboard

```{r}
#setting the working directory
setwd('~/your/working/directory')

#reading the data as a .csv file
diamond<-read.csv("diamonds.csv")

```

b). Remove the “depth” column and rename the columns “x”, “y” and “z” as “length”, “width” and “depth”.

```{r}
#Transforming the data into a data frame
diamond<-as.data.frame(diamond)

#Removing the depth column
diamond<-diamond[, -5]

#Renaming  the x,y and z columns
colnames(diamond)<-c("carat","cut","color","clarity","table","price",
                     "length","width","depth")
diamond

```
c).  The missing values are encoded differently in the data set, i.e. “missing”, “MISSING” and “NA”. Find the occurrence of each of them.

Values encoded under MISSING
```{r}
diamond[diamond=="MISSING"]

```

Values encoded under missing
```{r}
diamond[diamond=="missing"]

```

Values encoded under "NA"
```{r}
diamond[diamond=="NA"]
```


d). Assign NA to all missing values found above

```{r}
diamond[diamond=="MISSING"]<-NA
diamond[diamond=="missing"]<-NA
diamond[diamond=="NA"]<-NA
diamond
```
e). Find proportion of missing values per column
```{r}
colSums(is.na(diamond))/nrow(diamond)
```

f). Remove all missing values from the data set and create a new data set data.complete.

```{r}
data.complete<-na.omit(diamond)
data.complete

```
g). Check the structure of the data set and convert them to the appropriate data types

```{r}
str(data.complete)
```
The table, price, length and depth columns are characters when they should be integers
```{r}
#converting the table, price, length and depth columns into integers
data.complete$length<-as.numeric(data.complete$length)
data.complete$depth<-as.numeric(data.complete$depth)

str(data.complete)

```


h). For the numeric columns, replace the missing values with the column mean

```{r}
#Finding the means

mean(data.complete$carat)
mean(na.omit(data.complete$length))
mean(data.complete$width)
mean(na.omit(data.complete$depth))
mean(data.complete$table)
mean(data.complete$price)
```
```{r}
#replacing the means in missing columns
data.complete$carat[is.na(data.complete$carat)]<-mean(data.complete$carat)
data.complete$length[is.na(data.complete$length)]<-mean(na.omit(data.complete$length))
data.complete$width[is.na(data.complete$width)]<-mean(data.complete$width)
data.complete$depth[is.na(data.complete$depth)]<-mean(na.omit(data.complete$depth))
data.complete$table[is.na(data.complete$table)]<-mean(data.complete$table)
data.complete$price[is.na(data.complete$price)]<-mean(data.complete$price)
```

i). Make the cut, color and clarity columns into factors

```{r}
data.complete$cut<-as.factor(data.complete$cut)
data.complete$clarity<-as.factor(data.complete$clarity)
data.complete$color<-as.factor(data.complete$color)
str(data.complete)
```

j).Reorder the data frame by carat variable in the descending order and output the first 6 rows

```{r}
data.complete<-data.complete[order(-data.complete$carat), ]
head(data.complete)
```

k). Take a subset of the data frame so that diamonds with at least 0.2 carat, I color or above (D is the highest colorless diamond grade), VVS1 or VVS2 clarity, price between $330 to $400 are kept. How many diamonds in the data set satisfies this condition?

```{r}
#find color levels
data.complete$color

sub_set<-subset(data.complete,carat>=0.2 & clarity %in% c("VVS1","VVS2") &
                  price>=330 & price<=400 & color %in% c("I","J"))
sub_set
nrow(sub_set)
```

